# IE - Master in Business Analytics and Big Data
Recommendation Systems

Building a Non Personalized Recommendation Engine

Author: Rahul Singh

========================================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Initialization

Import libraries

```{r}
library(plyr)
library(dplyr)
library(tibble)
library(stats)
library(stringr) #to clean movie lens dataset; white-space etc
library(tidyr) #for extract function used in data cleaning in Movielens
library(lubridate) #to clean tags dataset in Movielens
library(reshape2)
library(ggplot2)
```

Load the critics dataset.

```{r}
rm(list=ls()) #to remove previously existing data in the R memory
critics<-read.csv("~/Desktop/Build Collaborative Recommendation Engine/IE-Lab1-Package/Data/critics.csv",check.names=FALSE) #inputing dataset in data frame
```

# 1) Top 5 Movies, by rating mean
Calculate mean rating for each movie, ordered with the highest rating listed first, and plot the top 5.

**Expected result**:
3.600000, 'The Shawshank Redemption'
3.266667, 'Star Wars IV - A New Hope'
3.222222, 'Blade Runner'
3.166667, 'Groundhog Day'
3.062500, 'The Silence of the Lambs'


```{r fig.width=7, fig.height=6}
#dropping the user name column
critics<-critics %>% select(-User) 

#calculating mean rating for each movie
mean_rating<- critics %>%  summarise_at(vars(colnames(.)),funs(mean(., na.rm=TRUE))) 

#saving column names
col=as.data.frame(colnames(mean_rating)) 

#transposing mean_rating and storing it as data frame
t_mean_rating<-as.data.frame(t(mean_rating))

#binding the transposed mean ratings and column names
final_mean_ratings<-cbind(t_mean_rating,col)
final_mean_ratings<-as.data.frame(final_mean_ratings)

#renaming the final mean ratings dataframe
colnames(final_mean_ratings)[1] <- "Avg_Rating"
colnames(final_mean_ratings)[2] <- "Movie"

#arranging the ratings in descending order and printing out the head with n=5 of result
result<- arrange(final_mean_ratings,desc(Avg_Rating))
result<-head(result, n=5)
print(result)
```

# 2) Top 5 Movies, by rating distribution
Calculate the percentage of ratings for each movie that are 4 stars or higher. Order with the highest percentage first, and plot the top 5 movies with more high ratings distribution.

**Expected result**:
0.7000000, 'The Shawshank Redemption'
0.5333333, 'Star Wars IV - A New Hope'
0.5000000, 'Gladiator'
0.4444444, 'Blade Runner'
0.4375000, 'The Silence of the Lambs'

```{r fig.width=7, fig.height=6}
#we have created new_data so that in case we want to transform the data to binary matrix in Q4 
#we don' change the original dataset 'critics'
new_data<- critics 

a<-summary(new_data>3) #to view roughly how many values are 4s and 5s; and NA

#to have a transposed dataframe called 'counting_frequencies' 
#and converting values to factores with five levels using apply function
counting_frequencies<-as.data.frame(t(apply(new_data, 2, function(x) table(factor(x, levels=1:5)))))

#counting frequencies in the new data.frame using rowSums and storing the values in a new columns called 'quantity_of_ratings'
counting_frequencies$quantity_of_ratings <- rowSums(counting_frequencies[1:5])

#getting frequencies for 4 or 5 and storing the number in a new column 'rating_4_or_5'
counting_frequencies$rating_4_or_5 <- rowSums(counting_frequencies[4:5])

#creating a new column called 'high_rating_distribution' and computing the frequencies
counting_frequencies$high_rating_distribution <- counting_frequencies[7]/counting_frequencies[6]

#now we have counting_frequencies with new columns of interest stored as a dataframe matrix
counting_frequencies<-as.data.frame.matrix(counting_frequencies)

#see the data frame created so far, we print it
print(head(counting_frequencies, n=5))

```

```{r}
#we need to unlist the newly created column called 'high_rating_distribution'
#we take the previous data frame matrix 'counting_frequencies' and convert it to a new dataframe called 'counting_frequencies_1'
#the values are stored as numeric
counting_frequencies_1<-transform(counting_frequencies, high_rating_distribution = as.numeric(unlist(high_rating_distribution)))

#order the data frame counting_frequencies_1 by high_rating_distribution column 
#store the results in a new dataframe called counting_frequencies_2
counting_frequencies_2<-counting_frequencies_1[order(counting_frequencies_1$high_rating_distribution, decreasing = TRUE),]

#adding column name "Movie" and storing the results in a new dataframe called counting_frequencies_3
counting_frequencies_3<- rownames_to_column(counting_frequencies_2, "Movie")

#Storing the values in a new data frame called counting_frequencies_4 and excluding columns 2 to 8
#This is the final result
counting_frequencies_4<- counting_frequencies_3[-c(2:8)]
print(head(counting_frequencies_4,n=5))
```

# 3) Top 5 Movies, by quantity of ratings
Count the number of ratings for each movie, order the one with most number of ratings first, submit the top 5.

**Expected result**:
17, 'Toy Story'
16, 'The Silence of the Lambs'
15, 'Star Wars IV - A New Hope'
14, 'Star Wars VI - Return of the Jedi'
13, 'Independence Day'

```{r fig.width=7, fig.height=6}
#we can use the counting_frequencies_3 dataframe previously created in question 2 to answer this question
#let's print the head of 
print(head(counting_frequencies_3,n=5)) #notice we are interested in column quantity_of_ratings

#order counting_frequencies_3 according to decreasing order in column called quantity_of_ratings
#store results in a new data frame called 'quantity'
quantity<-counting_frequencies_3[order(counting_frequencies_3$quantity_of_ratings, decreasing = TRUE),]

#remove columns that are not needed to get the final result
quantity_1<- quantity[-c(2:6,8,9)]

#we are putting the row.names to null in order to remove duplication in a column which already has the row names in it
row.names(quantity_1)<-NULL

#printing the final results for top 5
print(head(quantity_1,n=5))
```

# 4) People who watched Star Wars IV also watched ...
Calculate movies that most often occur with other movie. For each movie, calculate the percentage of the other movie raters who also rated that movie. Order with the highest percentage first, and submit the top 5. Select "Star.Wars.IV...A.New.Hope" movie by defaut. Use the (x+y)/x method.

Top Occurences:
0.9333333, 'Toy Story'
0.8666667, 'Star Wars: Episode VI - Return of the Jedi'
0.8000000, 'The Silence of the Lambs'
0.7333333, 'Independence Day'
0.6666667, 'Total Recall'

```{r fig.width=7, fig.height=6}
#we defined the critics dataset as new_data in Q2 and now we will use this data; 
#notice this does not have names of users; we removed it in Q1
#we want to convert new_data to a binary matrix by replacing NA values with 0 and positions where ratings are given with 1
new_data[new_data!=0] <-  1
new_data[is.na(new_data)] <- 0

#we call this data frame as 'binary_dataframe'
binary_dataframe<-new_data

#storing the column sums to a vector called sums
sums <- colSums(binary_dataframe)

#converting the binary_dataframe to 'as.data.frame' to ensure correct types of values in binary_dataframe
df<-as.data.frame(binary_dataframe)

#initializing n empty matrix where we will store the results
A <- data.frame(matrix(ncol = ncol(df), nrow = nrow(df)))

#to obtain values using count(x and y)/ count(x)
#first row represent relative association with respect to first movie
for (j in 1:ncol(df)) {
  for (i in 1:ncol(df)) { 
    A[i,j] = (length(which(df[i] == df[j] & df[i]!=0 & df[j]!=0)))/ sum(df[i])
  }
}
  
# Vector 'col' has the movie names stored in it
#we subset the first row of matrix A that has our results in vector called 'results'
results=rbind(colnames(binary_dataframe),A[1,])
results=t(results)
```

```{r}
#unlisting the results list and storing it as dataframe df
df <- data.frame(matrix(unlist(results), nrow=20,ncol=2, byrow=F),stringsAsFactors=FALSE)

#sorting the dataframe by descending order
df_1 <- arrange(df,desc(X2))

#subsetting the top results from df_1 to df_2
df_2<-df_1[2:6,c(1,2)]

#putting the rownames as null
rownames(df_2) <- NULL

#printing the top 5 results excluding the first row because it is the "Star.Wars.IV...A.New.Hope" movie itself
print(df_2,row.names=FALSE)
```

# 5) People who liked Babe also liked ...
Calculate the movies better rated of people who liked a movie. Select the people who liked the movie "Babe" (4 or 5 stars) and provide the top 5 movies they liked most.

**Expected result**:
5.000, 'Pulp.Fiction'
5.000, 'Groundhog.Day'
4.500, 'The.Shawshank.Redemption'
4.333, 'Toy.Story'
4.000, 'Blade.Runner'

```{r}
new_list <- list()
print_list<- list()

for(j in 1:ncol(critics)){
  
  new_list[[paste0("code_",j,sep="")]]<-colSums((critics %>% filter(.[j]>=4) ), na.rm=TRUE)/colSums(!is.na((critics %>% filter(.[j]>=4) )))
  
  new_list[[paste0("code_",j,sep="")]]<-new_list[[paste0("code_",j,sep="")]][-j] #removing the movie itself eg Babe vs Babe
  
  print_list[[paste0("print_",j,sep="")]] = new_list[[paste0("code_",j,sep="")]][order(new_list[[paste0("code_",j,sep="")]], decreasing=TRUE)[1:5]] #creating a list of lists with sub-list 20 is for Babe
  
}

#print sub-list 20 for Babe
print(print_list[[20]])
```


# 6) MovieLens
Explore a real data set and provide non-personalized ratings. You can use the MovieLens database. You can find MovieLens' dataset here: http://files.grouplens.org/datasets/movielens/ml-10m-README.html

```{r}

#small dataset of Movielens dataset was downloaded and stored in the following variables
links<-read.csv("~/Desktop/Build Collaborative Recommendation Engine/IE-Lab1-Package/Building a Non Personalized Recommendation/MovieLens Data/ml-latest-small/links.csv")

movies<-read.csv("~/Desktop/Build Collaborative Recommendation Engine/IE-Lab1-Package/Building a Non Personalized Recommendation/MovieLens Data/ml-latest-small/movies.csv", stringsAsFactors=FALSE)

ratings<-read.csv("~/Desktop/Build Collaborative Recommendation Engine/IE-Lab1-Package/Building a Non Personalized Recommendation/MovieLens Data/ml-latest-small/ratings.csv")

tags<-read.csv("~/Desktop/Build Collaborative Recommendation Engine/IE-Lab1-Package/Building a Non Personalized Recommendation/MovieLens Data/ml-latest-small/tags.csv")


```

Printing the structure of Movielens dataset to observe its structure
```{r}
str(links)
```
```{r}
str(movies)
```
```{r}
str(ratings)
```

```{r}
str(tags)
```
```{r}
#changing the timestamp value in tags using lubridate function called as_datetime
tags_dataframe <- tags %>% mutate(timestamp = as_datetime(timestamp))
```

```{r}
#(1) we trim the white spaces with str_trim() function; (2) split the title to title and year variable;
#(3) we take out the debut date for the movie; (4) replace NA with the name of the title; (5) drop the timestamp column;
#(6) we need to put NA to those values which do not have genres

movies_dataframe <- movies %>% mutate(title = str_trim(title)) %>% extract(title, c("title_tmp", "year"), regex = "^(.*) \\(([0-9 \\-]*)\\)$", remove = F) %>% mutate(year = if_else(str_length(year) > 4, as.integer(str_split(year, "-", simplify = T)[1]), as.integer(year))) %>% mutate(title = if_else(is.na(title_tmp), title, title_tmp)) %>% select(-title_tmp)  %>% mutate(genres = if_else(genres == "(no genres listed)", `is.na<-`(genres), genres))
```

```{r}
# Number of movies per year/decade
movies_every_year <- movies_dataframe %>%
  na.omit() %>% # omit missing values
  select(movieId, year) %>% # select columns we need
  group_by(year) %>% # group by year
  summarise(count = n())  %>% # count movies per year
  arrange(year)
```

```{r}
# plot distribution of ratings
ggplot(data=ratings, aes(x=rating), aplha=0.8) + geom_histogram() + xlab('Ratings') + ylab('Counts')
```

```{r}
# fill missing years
movies_every_year <- movies_every_year %>% complete(year = full_seq(year, 1), fill = list(count = 0)) 
#we plot the number of movies for every year and we can observe a clear increase in the movies
movies_every_year %>% ggplot(aes(x = year, y = count)) + geom_line(color="brown") 

```

```{r}
#we save the genres which are multiple in number and are separated by |
#we store the results in genres_dataframe
genres_dataframe <- movies_dataframe %>% separate_rows(genres, sep = "\\|") %>% group_by(genres) %>% summarise(number = n()) %>% arrange(desc(number))
```

```{r}
#we change the timestamp column to proper readable format
ratings_dataframe <- ratings %>% mutate(timestamp = as_datetime(timestamp))

#we join ratings_dataframe to movie_dataframe with movieID as key and omit na rows; then we select four columns of our interest
average_rating <- ratings_dataframe %>% inner_join(movies_dataframe, by = "movieId") %>% na.omit() %>% select(movieId, title, rating, year) %>% group_by(movieId, title, year) %>% summarise(count = n(), mean = mean(rating), min = min(rating), max = max(rating)) %>% ungroup() %>% arrange(desc(mean))

print(head(average_rating, n=20))
```

```{r}
#create a function to comput the weighted rating for non personalized recommendation 
# RATING = average for the movie mean 
# VOTES_COUNT = count of votes for the movie
# MINIMUM_NEEDED = minimum votes required to be listed in the Top 250
# MEAN_ENTIRE = the mean vote across the entire dataset
weighted_rating_for_non_personalized <- function(RATING, VOTES_COUNT, MINIMUM_NEEDED, MEAN_ENTIRE) {
  return (VOTES_COUNT/(VOTES_COUNT+MINIMUM_NEEDED))*RATING + (MINIMUM_NEEDED/(VOTES_COUNT+MINIMUM_NEEDED))*MEAN_ENTIRE
}

#apply this function to obtain average_rating
average_rating <- average_rating %>% mutate(wr = weighted_rating_for_non_personalized(mean, count, 500, mean(mean))) %>% arrange(desc(wr))
print(head(average_rating, n=20))
```






