# IE - Master in Business Analytics and Big Data
Recommendation Systems

Building a Collaborative Filtering Engine

Author: Rahul Singh

========================================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initialization

Import libraries

```{r}
library(plyr)
library(stats)
library(ggpubr) #to plot scatter plots between Pearson Coefficients of two user we need this package
```

Load the critics dataset.

```{r}
rm(list=ls()) #to remove previously existing data in the R memory

#inputing dataset in data frame
critics<-read.csv("~/Desktop/Build Collaborative Recommendation Engine/IE-Lab1-Package/Data/critics.csv",check.names=FALSE) 
```

# 1) Pearson correlation coefficient 
Calculate the Pearson correlation coefficient between Victoria and Nuria

**Expected correlation between Victoria and Nuria**: 0.3651484

```{r fig.width=7, fig.height=6}
critics_t <- data.frame(t(critics)) #transpose the dataframe critics

#sane the transposed critics_t as character because after transposing its format gets changed
#user names are also not seen in the transposed critics_t
critics_t[] <- lapply(critics_t, as.character)

#extract the row names from critics 1st row and save it as colnames of transposed dataframe called critics_t
colnames(critics_t) <- critics_t[1, ]

#drop the name of the movies in critics_t first column
critics_t <- critics_t[-1 ,]

#convert the values back to numeric from character
critics_t[] <- lapply(critics_t, function(x) as.numeric(as.character(x))) #now our dataframe is numeric

#correlation of Victoria and Nuria
print(cor(critics_t$Victoria,critics_t$Nuria, use = "complete.obs")) #correlation of Victoria and Nuria
```

# 2) Compare two movie critics
Using the calculated Pearson correlation in the previous section. Compare and show the movie ratings for Victoria and Nuria on a scatter plot. How is this compared with the same ratings of Maria and Nerea on a scatter plot? What about Chris and Jim? What it means?

```{r}
#ggscatter function with cor.method as "pearson" does a scatter plot for the movie ratings for Victoria and Nuria
#notice that in x axis we are plotting Victoria's and y axis Nuria's ratings
#grey area also shows the confidence interval with a regression line

#the relation is low; this means that their ratings are not that similar

ggscatter(critics_t, x = "Victoria", y = "Nuria",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Victoria", ylab = "Nuria")
```

```{r}
#we do the same as above and observe that Maria and Nerea have negative correlation
ggscatter(critics_t, x = "Maria", y = "Nerea",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Maria", ylab = "Nerea")
```

```{r}
#Chris and Jim have a higher positive correlation; this shows that their movie ratings have a similarity of nearly 83 percent

ggscatter(critics_t, x = "Chris", y = "Jim",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Chris", ylab = "Jim")
```


# 3) Top Recommendations 
Get the top 5 movies recommended for Victoria. by using a weighted average of every other critic's rankings.

**Expected result**:
'The Matrix'
'Forrest Gump'     
'The Sixth Sense' 
'Shakespeare in Love'
'Blade Runner'

```{r}
correlation_matrix=cor(critics_t, use = "pairwise.complete.obs")
row_names= row.names(critics_t)
col_names= colnames(critics_t)

correlation_Victoria_with_others= correlation_matrix[,15]
rows_to_select<- which(is.na(critics_t["Victoria"])) #gives location of NA 
#and we can look at who else has rated the movie which Victoria has not rated

matrix_1= matrix(, nrow = 20, ncol = 20)

j=1
for (j in rows_to_select){ 
  #j only selects those movies for which Victoria has not rated
  for (i in 1:20) {
      if (i!=15) { #i!=15 helps not select Victoria's column
        #in this we calculate the difference of user's rating in each cell 
        # (contd.) minus his/her average user rating times Pearson coef/sum of absolute values of Pearson coeff
      matrix_1[i,j] <- ((critics_t[j,i]- colMeans(critics_t[i], na.rm=T)) * correlation_Victoria_with_others[i])/sum(abs(correlation_Victoria_with_others)) 
      }
    }
}

```

```{r}
new_list=colSums(matrix_1, na.rm=T)  #new_list which has sum for each movie's weighted average

#now we are missing adding Victoria's rating to the sum of values to get the predicted score for each movie that Victoria has not rated

#note that zero in the list represent those movie's index for which Victoria has rated

for (i in 1:length(new_list)) {
  if (new_list[i] ==0) {
    new_list[i]<- NA
  }
}

Victoria_average_rating=colMeans(critics_t[15], na.rm=T)

#adding Victoria's average rating to the values
new_list<-new_list+Victoria_average_rating

#to print the values with positions which represent the indexes of the movies in the critics_t (transposed matrix)
print(new_list) 
```

```{r}
#order the new_list in decreasing order and extract the movie index
a<-order(new_list, decreasing=TRUE)


#select the top 5 movies recommended for Victoria. by using a weighted average of every other critic's rankings
a<-a[1:5]

print(row.names(critics_t[a,]))
```




# 4) Top similar critics
Return the top 5 critics that are most similar to Victoria. 

**Expected result**:
0.9449112, 'Rachel'
0.5976143, 'Ana' 
0.5789794, 'Oriol'
0.4925922, 'Maria'
0.4273247, 'Carles'

```{r fig.width=7, fig.height=6}

top_results<-as.data.frame(correlation_Victoria_with_others)
top_results$names <- rownames(top_results)
rownames(top_results)<-NULL
final<-top_results[order(desc(top_results$correlation_Victoria_with_others)),]

#drop first row as it contans Victoria's correlation with herself which is 1
final<-final[-1,]
rownames(final)<-NULL
print(head(final,n=5))
```